On the characteristic and diameter of planar integral point sets \footnote{ The work of the first author was partially supported by the Theoretical Physics and Mathematics Advancement Foundation “BASIS”. }

### 2024-07-11 char-4k-bound/char-4k-bound.tex
[4k3] Embrace some more fixes from Dr. A.S. Usachev

\begin{definition}
	The characteristic of a planar IPS $\mathcal{M}$ is a squarefree number $q$
	such that the area of any triangle $M_1M_2M_3$, $\{M_1,M_2,M_3\}\subset M$,
	is commensurable with $\sqrt{q}$.
	Notation: $\chr M = q$.
\end{definition}

@@

\begin{proof}
	Let $|BC| = n$. Then, from the triangle inequality $|AC| < |AB| + |BC|$, we have that $|AC| = m + n - s$, $s \in \mathbb{N}$ and $s \leq m$. By the Grid Theorem, we introduce such coordinates that $A = (-\frac{m}{2}; 0)$, $B = (\frac{m}{2}; 0)$, and $C = (\frac{a}{2m}; \frac{b\sqrt{p}}{2m})$. We find the distance between points $A$ and $C$, and points $B$ and $C$ in coordinates, and form a system of equations
	\begin{equation}
		\label{eq:29}
		\begin{cases}
		\frac{\sqrt{(a+m^2)^2+pb^2}}{2m}&=m+n-s.\\
		\frac{\sqrt{(a-m^2)^2+pb^2}}{2m}&=n
		\end{cases}
	\end{equation}
	Multiply each equation by $2m$ and square both sides of the equations in system (\ref{eq:29})
	\begin{equation}
		\label{eq:30_pre_weeding_I}
		\begin{cases}
		(a+m^2)^2+pb^2&=4m^2(m+n-s)^2,\\
		(a-m^2)^2+pb^2&=4m^2n^2
		\end{cases}
	\end{equation}
	and then subtract the second equation from the first one:
	\begin{equation}
	(a+m^2)^2-(a-m^2)^2=4m^2(m+n-s)^2-4m^2n^2,
	\label{eq: 31}
	\end{equation}
	$$
	4m^2a=4m^2((m+n-s)^2-n^2),
	$$
	so
	\begin{equation}
		\label{eq:32_pre_weeding_I}
		a=(m-s)(m+2n-s).
	\end{equation}
	Substitute the obtained expression~\eqref{eq:32_pre_weeding_I}
	into the second equation of the system~\eqref{eq:30_pre_weeding_I}:
	\begin{equation}
		((m-s)(m+2n-s)-m^2)^2+pb^2=4m^2n^2,
	\end{equation}
	that immediately gives
	\begin{equation}
		\label{eq:33}
		pb^2=s(2m+2n-s)(2n-s)(2m-s).
	\end{equation}
	According to the assumption, let $s=2t+1$, where $t\in\mathbb{N}_0$.
	Then equation (\ref{eq:33}) can be rewritten as:
	\begin{equation}
		\label{eq:34.2}
		pb^2=(2t+1)(2m+2n-2t-1)(2n-2t-1)(2m-2t-1)
	\end{equation}
	or
	\begin{eqnarray}
		pb^2+1 &=&4(-4t^4+8t^3m+8t^3n-8t^3-4t^2m^2-12t^2mn+12t^2m- {}\nonumber\\
			 &     &-4t^2n^2+12t^2n-6t^2+4tm^2n-4tmn^2-12tmn+6tm- {}\nonumber\\
			 &     &-4tn^2+6tn-2t+2m^2n-m^2+2mn^2-3mn+m-n^2+n).%{}\nonumber\\
	\label{eq:35}
	\end{eqnarray}
	If equation (\ref{eq:35}) has integer solutions, then both $p$ and $b$ are odd.
	The right-hand side is divisible by 4. Then $pb^2\equiv3\operatorname{mod}{4}$.
	But $b^{2}\equiv1(\operatorname{mod}4)$.
	Thus, $p=4k+3$, $k\in \mathbb{N}_{0}$.
\end{proof}

Then the following theorem immediately appears.

@@

\begin{proof}
	Let $|M_1 M_2| = m$, $m \in \mathbb{N}$, and let the $n$-th Erdős curve satisfy $|M_1 M_3| - |M_2 M_3| = n$.
	If $m$ is an odd number, then by Lemma~\ref{lem:pre_weeding_I}, the difference $n = |M_1 M_3| - |M_2 M_3| = m - s$, where $s$ is an even number, is also an odd number.
	Let $m$ be an even number; then $n = |M_1 M_3| - |M_2 M_3| = m - s$ is an even number, as the difference of two even numbers.
\end{proof}

An even edge is a bit special case.

@@

\begin{proof}
	Let us set $m = 2q$ in the Lemma~\ref{lem:pre_weeding_I}.
	Then equation \eqref{eq:33} turns into
	\begin{equation}
		\label{eq:33.1}
		pb^2 = s(4q + 2n - s)(2n - s)(4q - s).
	\end{equation}
	Taking into account that $s = 2t + 1$, $t \in \mathbb{N}_0$,
	we bring equation (\ref{eq:33.1}) to the form:
	\begin{equation}
	pb^2 = (2t + 1)(4q + 2n - 2t - 1)(2n - 2t - 1)(4q - 2t - 1).
	\label{eq: 34.2}
	\end{equation}
	or, after expansion,
	\begin{eqnarray}
		pb^2 + 1&=& 8(-2t^4+8t^3q+4t^3n-4t^3-8t^2q^2-12t^2qn+12t^2q- {}\nonumber\\
			 &     &-2t^2n^2+6t^2n-3t^2+8tq^2n-4tqn^2-12tqn+6tq- {}\nonumber\\
			 &     &-2tn^2+3tn-t+4q^2n-2q^2+2qn^2-3qn+q)-  {}\nonumber\\
			   &     &-4n(n-1).
		\label{eq:35.1}
	\end{eqnarray}
	Among two consecutive natural numbers $n$ and $(n-1)$,
	exactly one of them is divisible by 2.
	Therefore, the expression $4n(n-1)\equiv0\pmod{8}$.
	Thus, the right-hand side of equation (\ref{eq:35.1}) is an even number and a multiple of 8.
	Then $pb^2\equiv7\pmod{8}$.
	A necessary condition for the existence of an integer solution to equation (\ref{eq:35.1})
	is that numbers $p$ and $b$ must be odd.
	But $b^2\equiv1\pmod{8}$, so $p=8k+7$, $k\in\mathbb{N}_0$.
\end{proof}

@@

	Assume the opposite. Let $M={M_1, M_2, M_3, M_4} \in \overline{\mathfrak{M}_4}'$. Introduce Cartesian coordinates such that $M_1(-1;0), M_2(1;0), O(0;0)$. Then $M_3$ and $M_4$ lie on the median perpendicular to $M_1M_2$. Since the distance $|M_3M_4|$ is an integer, the area of triangle $M_1M_3M_4$ is a rational number. By Definition 6, this means that $\operatorname{char}{M}={1}$. Then by the theorem about the lattice we have $M_3(0;t/4)$. Without loss of generality, let us assume $t\in\mathbb{N}$. Let $|M_1M_3|=s$. Applying the Pythagorean theorem to triangle $OM_1M_3$, we get
	$$
	1+\frac{t^2}{16}=s^2,
	$$
	or equivalently,
	\begin{equation}
	\label{eq: 42}
	16s^2-t^2=16.
	\end{equation}
	Find the solutions to equation (\ref{eq: 42}) in natural numbers. To do this, we introduce the substitution: $t=4k$, $k\in\mathbb{N}$. Then equation (\ref{eq: 42}) takes the form:
	\begin{equation}
	\label{eq: 43}
	s^2-k^2=1.
	\end{equation}
	Break down the left-hand side of equation (\ref{eq: 43}) into factors:
	\begin{equation}
	\label{eq: 44}
	(s-k)(s+k)=1.
	\end{equation}
	Since $s,k\in\mathbb{N}$, the number 1 can be represented as the product of two integers in two ways: $1\cdot1$ and $-1\cdot(-1)$. In the first case, $(s-k)=1$ and $(s+k)=1$, which gives $s=1,k=0$, which contradicts that $k\in\mathbb{N}$.
	In the second case, $(s-k)=-1$ and $(s+k)=-1$, which gives $s=-1,k=0$, which contradicts that $s,k\in\mathbb{N}$.

	Thus, equation (\ref{eq: 42}) has no natural solutions in natural numbers.
	The obtained contradiction completes the proof of the lemma.
\end{proof}

The following result is a key detail of bounds presented
in~\cite{solymosi2003note} and~\cite{my-semi-general-5-4-bound-2019}:

\begin{lemma}
	\cite[Observation 1]{solymosi2003note}
	If a triangle $T$ has integer side lengths $a \leq b \leq c$,
	then the minimal height $m$ of it is at least $\left(a - \frac{1}{4}\right)^{1/2}$.
\end{lemma}

If we exclude characteristics of the form $4k+3$, this Solymosi's result can be sharpened a bit.

@@

\begin{proof}
	Indeed, consider the triangle $ABC$ and let $|BC|=a$, $|AC|=b$.
	From the triangle inequality $|AB| < |BC|+|AC|$,
	we can represent the length of side $AB$ as $|AB|=a+b-s$, $s \in \mathbb{N}$ and $s \leq a$.
	For $s=1$, the conditions of Lemma~\ref{lem:pre_weeding_I} are satisfied: $s$ is an odd number and $s \leq a$.
	Therefore, the triangle with sides $a$, $b$, and $c=a+b-1$
	has a characteristic of the form $p=4k+3$, $k\in \mathbb{N}_0$.
\end{proof}

>>>>

\begin{definition}
	\label{def:char}
	The characteristic of a planar IPS $\mathcal{M}$ is a squarefree number $q$
	such that the area of any triangle $M_1M_2M_3$, $\{M_1,M_2,M_3\}\subset M$,
	is commensurable with $\sqrt{q}$.
	Notation: $\chr M = q$.
\end{definition}

@@

\begin{proof}
	Let $|BC| = n$. Then, from the triangle inequality $|AC| < |AB| + |BC|$, we have that $|AC| = m + n - s$, $s \in \mathbb{N}$ and $s \leq m$.
	By the Grid Theorem, we can assume that $A = (-\frac{m}{2}; 0)$, $B = (\frac{m}{2}; 0)$, and $C = (\frac{a}{2m}; \frac{b\sqrt{p}}{2m})$. We find the distance between points $A$ and $C$, and points $B$ and $C$ in coordinates, and form a system of equations
	\begin{equation}
		\label{eq:29}
		\begin{cases}
		\frac{\sqrt{(a+m^2)^2+pb^2}}{2m}&=m+n-s.\\
		\frac{\sqrt{(a-m^2)^2+pb^2}}{2m}&=n
		\end{cases}
	\end{equation}
	Multiply each equation by $2m$ and square both sides of the equations in system (\ref{eq:29})
	\begin{equation}
		\label{eq:30_pre_weeding_I}
		\begin{cases}
		(a+m^2)^2+pb^2&=4m^2(m+n-s)^2,\\
		(a-m^2)^2+pb^2&=4m^2n^2
		\end{cases}
	\end{equation}
	and then subtract the second equation from the first one:
	\begin{equation}
	(a+m^2)^2-(a-m^2)^2=4m^2(m+n-s)^2-4m^2n^2,
	\label{eq: 31}
	\end{equation}
	$$
	4m^2a=4m^2((m+n-s)^2-n^2),
	$$
	so
	\begin{equation}
		\label{eq:32_pre_weeding_I}
		a=(m-s)(m+2n-s).
	\end{equation}
	Substitute the obtained expression~\eqref{eq:32_pre_weeding_I}
	into the second equation of the system~\eqref{eq:30_pre_weeding_I}:
	\begin{equation}
		((m-s)(m+2n-s)-m^2)^2+pb^2=4m^2n^2,
	\end{equation}
	that immediately gives
	\begin{equation}
		\label{eq:33}
		pb^2=s(2m+2n-s)(2n-s)(2m-s).
	\end{equation}
	Since $s$ is assumed to be odd, let $s=2t+1$, where $t\in\mathbb{N}_0$.
	Then equation (\ref{eq:33}) can be rewritten as:
	\begin{equation}
		\label{eq:34.2}
		pb^2=(2t+1)(2m+2n-2t-1)(2n-2t-1)(2m-2t-1)
	\end{equation}
	or
	\begin{eqnarray}
		pb^2+1 &=&4(-4t^4+8t^3m+8t^3n-8t^3-4t^2m^2-12t^2mn+12t^2m- {}\nonumber\\
			 &     &-4t^2n^2+12t^2n-6t^2+4tm^2n-4tmn^2-12tmn+6tm- {}\nonumber\\
			 &     &-4tn^2+6tn-2t+2m^2n-m^2+2mn^2-3mn+m-n^2+n).%{}\nonumber\\
	\label{eq:35}
	\end{eqnarray}
	If equation (\ref{eq:35}) has integer solutions, then both $p$ and $b$ are odd.
	The right-hand side is divisible by 4, which implies that  $pb^2\equiv3\operatorname{mod}{4}$.
	Since $b^{2}\equiv1(\operatorname{mod}4)$, it follows that $p=4k+3$, $k\in \mathbb{N}_{0}$.
\end{proof}

The following theorem follows immediately.

@@

\begin{proof}
	Let $|M_1 M_2| = m$, $m \in \mathbb{N}$, and let the $n$-th Erdős curve satisfy $|M_1 M_3| - |M_2 M_3| = n$.
	If $m$ is an odd number, then by Lemma~\ref{lem:pre_weeding_I}, the difference $n = |M_1 M_3| - |M_2 M_3|$ equals $m - s$, where $s$ is an even number, is also an odd number.
	Let $m$ be an even number; then $n = |M_1 M_3| - |M_2 M_3| = m - s$ is an even number, as a difference of two even numbers.
\end{proof}

%An even edge is a bit special case.
The case when the length of an edge is even is slightly more specific and allows a more precise statement.

@@

\begin{proof}
	Let us set $m = 2q$ in the Lemma~\ref{lem:pre_weeding_I}.
	Then equation \eqref{eq:33} turns into
	\begin{equation}
		\label{eq:33.1}
		pb^2 = s(4q + 2n - s)(2n - s)(4q - s).
	\end{equation}
	Taking into account that $s = 2t + 1$, $t \in \mathbb{N}_0$,
	we bring equation (\ref{eq:33.1}) to the form:
	\begin{equation}
	pb^2 = (2t + 1)(4q + 2n - 2t - 1)(2n - 2t - 1)(4q - 2t - 1).
	\label{eq: 34.2}
	\end{equation}
	or, after expansion,
	\begin{eqnarray}
		pb^2 + 1&=& 8(-2t^4+8t^3q+4t^3n-4t^3-8t^2q^2-12t^2qn+12t^2q- {}\nonumber\\
			 &     &-2t^2n^2+6t^2n-3t^2+8tq^2n-4tqn^2-12tqn+6tq- {}\nonumber\\
			 &     &-2tn^2+3tn-t+4q^2n-2q^2+2qn^2-3qn+q)-  {}\nonumber\\
			   &     &-4n(n-1).
		\label{eq:35.1}
	\end{eqnarray}
	Among two consecutive natural numbers,
	exactly one is even.
	Therefore, the expression $4n(n-1)\equiv0\pmod{8}$.
	Thus, the right-hand side of equation (\ref{eq:35.1}) is a multiple of 8.
	Then $pb^2\equiv7\pmod{8}$.
	A necessary condition for the existence of an integer solution to equation (\ref{eq:35.1})
	is that numbers $p$ and $b$ must be odd.
	Since $b^2\equiv1\pmod{8}$, it follows that $p=8k+7$, $k\in\mathbb{N}_0$.
\end{proof}

@@

	Suppose, to the contrary, that $M={M_1, M_2, M_3, M_4} \in \overline{\mathfrak{M}_4}'$.
	Employ th Grid Theorem and set $M_1(-1;0), M_2(1;0), O(0;0)$.
	Then $M_3$ and $M_4$ lie on the perpendicular bisector to $M_1M_2$.
	Since the distance $|M_3M_4|$ is an integer, the area of triangle $M_1M_3M_4$ is a rational number.
	By Definition~\ref{def:char}, this means that $\operatorname{char}{M}={1}$.
	Then by the Grid Theorem we have $M_3(0;t/4)$. Without loss of generality, let us assume that $t\in\mathbb{N}$. Let $|M_1M_3|=s$. Applying the Pythagorean theorem to triangle $OM_1M_3$, we get
	$$
	1+\frac{t^2}{16}=s^2,
	$$
	or equivalently,
	\begin{equation}
	\label{eq: 42}
	16s^2-t^2=16.
	\end{equation}
	Now, we find the solutions to equation (\ref{eq: 42}) in natural numbers.
	To do this, we introduce the substitution: $t=4k$, $k\in\mathbb{N}$.
	Then equation (\ref{eq: 42}) takes the form:
	\begin{equation}
	\label{eq: 43}
	s^2-k^2=1.
	\end{equation}
	Break down the left-hand side of equation (\ref{eq: 43}) into factors:
	\begin{equation}
	\label{eq: 44}
	(s-k)(s+k)=1.
	\end{equation}
	Since $s,k\in\mathbb{N}$, the number 1 can be represented as the product of two integers in two ways: $1\cdot1$ and $-1\cdot(-1)$. In the first case, $(s-k)=1$ and $(s+k)=1$, which gives $s=1,k=0$, which contradicts that $k\in\mathbb{N}$.
	In the second case, $(s-k)=-1$ and $(s+k)=-1$, which gives $s=-1$, $k=0$, which also contradicts that $s,k\in\mathbb{N}$.

	Thus, equation (\ref{eq: 42}) has no natural solutions.
	The obtained contradiction completes the proof.
\end{proof}

The following result is a key element for bounds presented
in~\cite{solymosi2003note} and~\cite{my-semi-general-5-4-bound-2019}:

\begin{lemma}
	\cite[Observation 1]{solymosi2003note}
	If a triangle $T$ has integer side lengths $a \leq b \leq c$,
	then its minimal height $m$ is at least $\left(a - \frac{1}{4}\right)^{1/2}$.
\end{lemma}

If we exclude characteristics of the form $4k+3$, this Solymosi's result can be sharpened slightly.

@@

\begin{proof}
	Indeed, consider the triangle $ABC$ and let $|BC|=a$, $|AC|=b$.
	Using the triangle inequality $|AB| < |BC|+|AC|$,
	we can represent the length of side $AB$ as $|AB|=a+b-s$, $s \in \mathbb{N}$ and $s \leq a$.
	For $s=1$, the conditions of Lemma~\ref{lem:pre_weeding_I} are satisfied: $s$ is an odd number and $s \leq a$.
	Therefore, the triangle with sides $a$, $b$, and $c=a+b-1$
	has a characteristic of the form $p=4k+3$, $k\in \mathbb{N}_0$.
\end{proof}
